### 2010-03-30 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: merge typos fix from Pedro

We shall now fulfil our promises for $\Desc$. We are going to
implement the codes of the universe of descriptions. Interestingly, in
and by itself, the system of codes, $\Desc$, is nothing but a data-type. We
are in the same situation than with $\EnumU$: we ought to be able to describe
the codes of $\Desc$ in $\Desc$ itself. Hence, this code would be a
first-class citizen, born with the standard, generic equipment of
data-types.

@@

Let us explain how we have proceeded and the obstacle we face. Much as
the data-types we have seen so far, we first have to choose a
constructor. Constructors correspond to the code's constructors:
$\DUnit$, $\DSigma{\!}{\!}$, and $\DIndx{\!}$. The reader will notice that the
tagged notation we have used so far for $\Desc$ constructors now fully
make sense: these were actually the tags we are defining. \note{This
  sentence badly needs a rephrasing.}

@@

Equipped with $\DHindx{\!}{\!}$, we can describe the extended code of
our universe of data-types, as shown in
Figure~\ref{fig:desc-levitate}.  The $\DUnit$ and $\DIndx{\!}$ cases
remains unchanged, as expected. We successfully describe the
$\DSigma{\!}{\!}$ case, by a simple appeal to the higher-order
induction on $\V{S}$. The $\DHindx{\!}{\!}$ case consists in packing
the $\V{H}$ in $\Set$ with a recursive argument.

>>>>

We shall now fulfil our promises for $\Desc$. We are going to
implement the codes of the universe of descriptions. In and by itself,
the system of codes, $\Desc$, is nothing but a data-type. We are in
the same situation as with $\EnumU$: we ought to be able to describe
the codes of $\Desc$ in $\Desc$ itself. Hence, this code would be a
first-class citizen, born with the standard, generic equipment of
data-types.

@@

Let us explain how we have proceeded and the obstacle we face. Much as
the data-types we have seen so far, we first have to choose a
constructor. Constructors correspond to the code's constructors:
$\DUnit$, $\DSigma{\!}{\!}$, and $\DIndx{\!}$. The reader will notice
that the tagged notation we have used for the $\Desc$ constructors now
fully makes sense: these were actually the tags we are
defining.

@@

Equipped with $\DHindx{\!}{\!}$, we can describe the extended code of
our universe of data-types, as shown in
Figure~\ref{fig:desc-levitate}.  The $\DUnit$ and $\DIndx{\!}$ cases
remain unchanged, as expected. We successfully describe the
$\DSigma{\!}{}$ case, by a simple appeal to the higher-order induction
on $\V{S}$. The $\DHindx{\!}{\!}$ case consists in packing the $\V{H}$
in $\Set$ with a recursive argument.

### 2010-03-30 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: merge typos fix from Pedro

So far, the examples we have seen lives in $\IDesc{\Unit}$, hence are
not using any indexing. We are going to define a syntax for a small
typed language. We consider two types, natural numbers and booleans:

@@

Instead of implementing $\substH$ in this special case, we shall
implement it in a generic setting, and express our example in that
setting. However, before entering the realm of generic structures, let
us further explore the universe of indexed descriptions.

\subsection{Constrained constructors}

\begin{wstructure}
!!! Need Help !!!
<- Assuming a suitable notion of definitional equality for I
    ???
\end{wstructure}

In the following, we ask $I$ to come equipped with a notion of
definitional equality. Hence, we can legibly equates elements of $I$.

\note{ I will need some tutorial on this ad-hoc notion of definitional
  equality. Or what Conor's meant with that. }

\begin{wstructure}
<- Fin [figure]
    <- Presentation with equality constraints
    <- Example of constrained constructor
        -> Constraints are translated into equalities
    ???
\end{wstructure}

Another common dependent data-type is $\Fin{\!}$, the finite sets of
numbers. $\Fin{\!}$ is a typical example of an inductive family using
\emph{constrained constructors}. The constructors $\FinZero$ and
$\FinSuc{\!}$ are only defined for an index distinct from $\NatZero$:

\[
\stk{
\data \Fin{\!} : \PI{\V{n}}{\Nat} \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \FinZero_{\V{n}}      & \Fin{(\NatSuc{\V{n}})}   \\
    \FinSuc{\!}_{\V{n}}   & \Fin{\V{n}} \To \Fin{(\NatSuc{\V{n}})}
\end{array}
}
\]

\[\stk{
\FinD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}ll}
\FinD\: \V{n} = \DSigma{\!}{\!} \!\!\!\!& (\EnumT{[ \FinZero , \FinSuc{\!} ]}) \\
                        & \bigRedBracket{
                          \begin{array}{l}
                            \DSigma{\Nat}{\LAM{\V{m}} \DConst{(\V{n} \PropEq \NatSuc{\V{m}})}} \\
                            \DSigma{\Nat}{\LAM{\V{m}} \DProd{\DVar{\V{m}}}{\DConst{(\V{n} \PropEq \NatSuc{\V{m}})}}}
                          \end{array}
                          }
\end{array}
}\]

Intuitively, the constraint is captured as follows. First, we store an
element $\V{m}$ of $\Nat$, thanks to a $\DSigma{\!}{\!}$. However, the
constraint stipulates that $\V{m}$ cannot be \emph{any} natural
numbers: it must be ``the index minus one''. This translates into the
constraint $\V{n} \PropEq \NatSuc{\V{m}}$.

\note{ Anything else? }

\begin{wstructure}
<- GADT-style data-type definition
    <- We directly support it
        <- Constraints turned into equalities
            <- Fin example
    <- Again, smooth transition from "today" data-types to advanced ones
        <- Same grammar
        /> More power
    /> Equations in data-types are not a blessing
        <- Complexify induction principle
        <- Lose match between data-type definition and the elimination form
\end{wstructure}

The $\Fin{\!}$ example is also a well-known example of Generalised
Abstract Data-Type (GADT) in Haskell. It comes as no surprise that our
universe of indexed description supports this restricted class of
inductive families. As we have seen with $\Fin{\!}$, constrained
constructors are straightforwardly translated into equations.

However, one could, quite rightfully, be worried by the introduction
of equalities in the picture. In a programming context, equalities
haunt us, begging for proofs. Hence, whereas the developer typed a
constructor-constrained data-type, the elimination form will involve
equalities and their proofs. As much as possible, one would rather
avoid suffering from this burden. However, the equalities introduced
here are rather special: they are \emph{equations} constraining
variables to certain terms. Goguen et
al.~\cite{goguen:pattern-matching} have shown, in the context of
dependent pattern matching, how to tame these equations.

Further, we can actually remove these equations altogether. Let us
look back at $\Fin{\!}$. We note that the equations are introduced because
we are \emph{storing} the index of the inductive family, through
$\V{m}$. Hence the constraints. However, \emph{inductive families need not
  store their indices}~\cite{brady:index-inductive-families}. Applying
the \emph{forcing} optimisation to our initial definition of $\Fin{\!}$
gives the following, equivalent definition:

\[\stk{
\FinD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}llll}
\FinD\: \NatZero     & = & \multicolumn{2}{l}{\DSigma{(\EnumT{[]})}{\Void} } \\
\FinD\: (\NatSuc{\V{n}}) & = & \DSigma{\!}{\!}\!\!\!\!\! & (\EnumT{[ \FinZero , \FinSuc{\!} ]}) \\
                     &   &             & \bigRedBracket{
                                         \begin{array}{l}
                                         \DConst{\Unit} \\
                                         \DVar{\V{n}}
                                         \end{array}
                                         }
\end{array}
}\]

Equations have simply disappeared. Note that it is straightforward to
turn this definition into a tagged form. We should precise that
forcing a description is not guaranteed to remove all constraints. It
is subject to future work to see if constraints can be entirely
eradicated, or presented more conveniently to the developer. Finally,
it is worth mentioning that the forcing optimisation is a
\emph{source-to-source} transformation of the description.

@@

In this section, we develop more structure for our indexed
data-types. In Section~\ref{sec:desc-free-monad}, we have constructed
a free monad operation on simple descriptions. Building on this
experience, we are going to present its equivalent in the indexed
world. The process the same. Namely, given an indexed functor, we
derive the indexed functor coding its free monad:
\note{pwm: Whoa there. Maybe we should say something about IMonads in
general before we get to this point?}

>>>>

We are going to define a syntax for a small typed language. We
consider two types, natural numbers and booleans:

@@

Instead of implementing $\substH$ in this special case, we can
implement it in a generic setting, a \emph{free indexed-monad construction}
that we will see in the next section.

@@

In this section, we develop more structure for our indexed
data-types. In Section~\ref{sec:desc-free-monad}, we have built a free
monad operation for simple descriptions. Building on this experience,
we are going to present its equivalent in the indexed world. The
process is the same. Namely, given an indexed functor, we derive the
indexed functor coding its free monad: \note{pwm: Whoa there. Maybe we
  should say something about IMonads in general before we get to this
  point?}
